A new method to explicitly estimate the shift of optimum along gradients in multispecies studies

### 2022-11-04 redaction/manuscript_PhD_version.Rmd
Create a version with Appendix included

---
title: "A new method to explicitly estimate the shift of optimum along gradients in multispecies studies"
subtitle: "Running title: Modelling of gradient optimum shifts"
classoption: a4paper
output:
  pdf_document:
    keep_tex: true
  word_document: default
  documentclass: article
fontsize: 12pt
header-includes:
- \usepackage{float}
- \usepackage{lscape}
- \usepackage{amsmath}
- \usepackage{caption}
- \usepackage{lineno}
- \linenumbers
#- \usepackage{setspace}\doublespacing
csl: journal-of-biogeography.csl
bibliography: Paper_EHMOS_cut.bib
---

@@

$^*$Corresponding author: Bastien Mourguiart, Université de Pau et des Pays de l'Adour, UFR des Sciences et Techniques de la côte basque, 1 allée du parc Montaury, 64600 Anglet, France. Email: bastien.mourguiart@univ-pau.fr

## Acknowledgments
We thank Damien Combrisson from the Ecrin National Park that contributed to earlier stages of this research. We also thank Alain Gueguen to allow us to use his data. B.M. was supported by the UPPA-E2S international chaire of Kerrie Mengersen.

## Conflict of interest statement
All authors declare no conflict of interest.

@@

## Data availability
The data and R scripts used to produce this manuscript will be available on GitHub and archived on Zenodo before publication.

@@

## Biosketch
**Bastien Mourguiart** main research focus is on modelling species distributions. He is particularly interested in quantifying and explaining species responses to environmental changes. This work is part of his Ph.D. work at the University of Pau. He and the other authors collaborate to study Orthoptera altitudinal distributions and their determinants.

Authors' contribution statement: B.M., B.L. and A.B. conceived the ideas for this paper; B.M. wrote the manuscript and conducted the analyses with advice from B.L., K.M. and A.B.; T.C., J.M., Y.B. and A.B. conceived the sampling design and Y.B. collected field data with support from T.C. and J.M.; All the authors read and approved the final manuscript.

>>>>

---
title: "A new method to explicitly estimate the shift of optimum along gradients in multispecies studies"
subtitle: "Running title: Modelling of gradient optimum shifts"
classoption: a4paper
geometry: "left=3.8cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output:
  pdf_document:
    keep_tex: true
  word_document: default
  documentclass: article
fontsize: 12pt
header-includes:
- \usepackage{float}
- \usepackage{lscape}
- \usepackage{amsmath}
- \usepackage{nopageno}
- \usepackage{caption}
- \usepackage{setspace}\setstretch{1.5}
csl: journal-of-biogeography.csl
bibliography: Paper_EHMOS_cut.bib
---

@@

@@

@@

\newpage
## Appendix S1: Description of EHMOS' parameters

In the occupancy model of EHMOS (see Eq. (3) in the main text), the four parameters ($\alpha$, $\theta$, $\delta$ and $\tau$) represent ecological descriptors of the species-environment relationship (Fig. S1):

* $\alpha$ corresponds to the maximum probability of occurrence, $\psi_{max}$, on the logit scale. Indeed, the maximum probability of occurrence is reached when gradient value is equal to the optimum, i.e. we have $\psi_j = \psi_{max}$ when $X_j = \theta$. Thus, Eq. (3) in the main text becomes:

\begin{equation}
  log(\frac{\psi_{max}}{1-\psi_{max}}) = \alpha \iff \psi_{max}  =   \frac{1}{1+ \exp(-\alpha)};
\end{equation}

* $\theta$ is the species optimum, the environmental value at which species reached its maximum probability of occupancy;

* $\delta$ is the species shift between two optima;

* $\tau$ represents the environmental tolerance of a species, i.e. the gradient range that a species can occupy. It can be related to the width, $\omega$, of the species response curve at a specified occupancy probability threshold, $p_{\omega}$, by:

\begin{equation}
  \omega = 2 \times \tau \sqrt{2 \times (\alpha - logit(p_{\omega}))} \iff \tau = 0.5 \times  \frac{\omega}{\sqrt{2 \times (\alpha - logit(p_{\omega}))}}
\end{equation}
with a $p_{\omega}$ usually specified at 0.05 to estimate the suitable range of a species (Michaelis & Diekmann, 2017).

Those four EHMOS parameters can also be derived from cGLMM coefficient estimates (Jamil et al., 2014):

\begin{align*}
  \alpha &= \beta_0 - \frac{{\beta_1}^2}{4 \beta_2} \\
  \theta &= - \frac{\beta_1}{2 \beta_2} \\
  \delta &= - \frac{\beta_1}{2 \beta_2} - (-\frac{\beta_1 + \beta_4}{2 (\beta_2 + \beta_5)}) \\
  \tau &= \sqrt{-\frac{1}{2 \beta_2}}
\end{align*}

```{r schema_param, fig.cap = "", fig.align='center'}

img <- png::readPNG(here::here("results", "figs", "schema_parameters.png"))
grid::grid.raster(img)

Figure S1.1: Schematic representation of ecological parameters that describes species-environment relationship.

### References

Jamil T., Kruk C., & Braak C.J.F. ter (2014) A Unimodal Species Response Model Relating Traits to Environment with Application to Phytoplankton Communities. _PLoS ONE_, **9**, e97583.

Michaelis J. & Diekmann M.R. (2017) Biased niches – Species response curves and niche attributes from Huisman-Olff-Fresco models change with differing species prevalence and frequency. _PLOS ONE_, **12**, e0183152.

\newpage
## Appendix S2: Description of performance metrics used in the simulation study

The four performance metrics used (bias, RMSE, interval coverage and interval score) are computed for each model at different levels (species, scenario, overall). First, for each species $i$ included in a particular scenario $s$, we computed average perfomance metrics over the thirty replications:

* Bias of species $i$ in scenario $s$ based on method $m$ is defined as the average of differences between absolute^[Absolute values are took because true shifts could either be positive or negative, thus a negative bias always represents an underestimation of the true shif magnitude whatever its direction.] estimates based on method $m$ and absolute true shift over the thirty replications:

$$Bias_{i,s,m} = \frac{1}{R} \sum_{r=1}^{R}|\hat{\delta}_{i,s,m,r}| - |\delta_{i,s}| \text{ ;}$$

* The RMSE at species-specific level is defined by:

$$RMSE_{i,s,m} = \sqrt{\frac{1}{R}\sum_{r=1}^{R} (\hat{\delta}_{i,s,m} - \delta_{i,s})^2} \text{ ;}$$

* The interval coverage for a species in a particular scenario is the ratio between the number of replciations in which confidence interval estimate contain the true shift and the total number of replications;

* The interval score of species $i$ in scenario $s$ based on method $m$ is defined as the average of interval scores obtained at each replication $r$. At the replication level, interval score is computed as follows:

\begin{equation*}
IS_{i,s,m,r}(l_{i,s,m,r}, u_{i,s,m,r}; \delta_{i,s}) = 
\begin{cases}
(u_{i,s,m,r}-l_{i,s,m,r}) + \frac{2}{\alpha}(l_{i,s,m,r} - \delta_{i,s}) &\text{if $\delta_{i,s} < l_{i,s,m,r}$}\\
(u_{i,s,m,r}-l_{i,s,m,r}) &\text{if $u_{i,s,m,r} > \delta_{i,s} > l_{i,s,m,r}$}\\
(u_{i,s,m,r}-l_{i,s,m,r}) + \frac{2}{\alpha}(\delta_{i,s} - u_{i,s,m,r}) &\text{if $\delta_{i,s} > u_{i,s,m,r}$}
\end{cases}
\end{equation*}

where $\alpha$ is the confidence level of the interval, here set at 0.05, $l_{i,s,m,r}$ and $u_{i,s,m,r}$ are the lower and upper limits of the interval estimated by model $m$ for species $i$ in scenario $s$ and replicate $r$.

At the scenario level, metrics are averaged over the twenty species:
$$\overline{Bias}_{s,m} = \frac{1}{N} \sum_{i=1}^{N} Bias_{i,s,m}$$
$$\overline{RMSE}_{s,m} = \frac{1}{N} \sum_{i=1}^{N} RMSE_{i,s,m}$$
$$\overline{Coverage}_{s,m} = \frac{1}{N} \sum_{i=1}^{N} Coverage_{i,s,m}$$
$$\overline{IS}_{s,m} = \frac{1}{N} \frac{1}{R} \sum_{i=1}^{N} \sum_{r=1}^{R} IS(\hat{l}_{i,s,m,r}, \hat{u}_{i,s,m,r}; \delta_{i,s})$$

## Appendix S3: Complementary results of the simulation study

```{r setup, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# set chunk default options for the document
knitr::opts_chunk$set(echo=FALSE, 
                      message=FALSE, 
                      warning=FALSE, 
                      fig.pos = 'H',
                      fig.align = 'center')

library(knitr)
library(bookdown)
source(here::here("make.R"))
```

\setcounter{table}{0}  
\def\tablename{Table S3.}

```{r convergence_cglmm}
tabS1 %>%
  dplyr::bind_cols("pos" = c(5:8, 1:4)) %>%
  dplyr::arrange(pos) %>%
  dplyr::select(-pos) %>%
  kableExtra::kable(format = "latex",
                    caption = "Number of simulation replications per scenario that at least one model parameter had an R-hat higher than 1.1 after the maximum of iterations was reached", booktabs = TRUE) %>%
  kableExtra::kable_styling(latex_options = "HOLD_position") %>%
  kableExtra::pack_rows("Uniform sampling", 1, 4) %>%
  kableExtra::pack_rows("Unbalanced sampling", 5, 8)
```

```{r computation_time}
tabS2 %>%
  dplyr::bind_cols("pos" = c(5:8, 1:4)) %>%
  dplyr::arrange(pos) %>%
  dplyr::select(-pos) %>%
  kableExtra::kable(format = "latex",
                    caption = "Average computation times in hours.",
                    booktabs = TRUE,
                    row.names = NA) %>%
  kableExtra::kable_styling(latex_options = "HOLD_position") %>%
  kableExtra::pack_rows("Uniform sampling", 1, 4) %>%
  kableExtra::pack_rows("Unbalanced sampling", 5, 8)
```

```{r}
tabS3 %>%
  dplyr::bind_cols("pos" = rep(c(5:8, 1:4), 4),
                   "metric" = rep(c("bias", "cov", "is", "rmse"), each = 8)) %>%
  dplyr::arrange(metric, pos) %>%
  dplyr::select(-c(pos, metric)) %>%
  kableExtra::kable(format = "latex",
                    caption = "Average of species specific performance metrics for the eight simulated scenarios after 30 replications. Numbers in brackets represent standard deviations. ",
                    booktabs = TRUE,
                    longtable = TRUE) %>%
  kableExtra::kable_styling(latex_options = "HOLD_position") %>%
  kableExtra::pack_rows("Bias", 1, 8) %>%
  kableExtra::pack_rows("Coverage", 9, 16) %>%
  kableExtra::pack_rows("IS", 17, 24) %>%
  kableExtra::pack_rows("RMSE", 25, 32)
```

```{r esm_graph_IS, fig.dim = c(8.3, 11.7), fig.align = 'center', fig.cap=""}

gridExtra::grid.arrange(g.IS, g.IS2, nrow = 2)

Figure S3.1: Distribution of species specific interval scores (IS) for the eight simulated scenarios after 30 replications, and for the three methods (top graph) or without cGLMM (bottom graph). Colored points represent average IS in species group depending on their optimum position (placed in the edge or in the middle of sampling range) along the simulated gradient, and their ecological specialization type relative to the width of their ecological niche (generalist species with broad ecological niche or specialist species with narrower ecological niche).

\newpage
\setcounter{table}{0}  
\def\tablename{Table S4.}

## Appendix S4: Departure from symmetric assumption

Models used in the manuscript make the assumption that species have symmetric unimodal relationships with environmental gradient (i.e. occupancy probabilities decrease at the same rate on both side of optimum, see Fig. S4.1). Species-environment relationships could however have other forms [@oksanen2002; @jansen2013]. For instance, species response curve (SRC) could be unimodal but skewed to one part of the gradient (i.e. asymmetric). Here we propose to investigate the effect of departure from models' assumption of symmetric unimodal species-gradient relationships on models accuracy by conducting a simulation study.

```{r, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

source(here::here("analysis/review01_simu_assymetric_curves.R"))
source(here::here("analysis/review05_make-figs-tabs-results_asymSRC.R"))
```

```{r figure type of SRC, fig.cap = "", fig.align='center'}
x <- 1000:3000
y.sym <- src_asym(x = x, opt = 2000, sd_l = 100, sd_r = 100, max = 0.9) # sd chose to simulate a width around 450
y.asym_l <- src_asym(x = x, opt = 2000, sd_l = 150, sd_r = 100, max = 0.9) 
y.asym_r <- src_asym(x = x, opt = 2000, sd_l = 100, sd_r = 150, max = 0.9)

# Figure: Representation of the three types of species response curves. 
data.frame("x" = x,
           "y" = c(y.sym, y.asym_l, y.asym_r),
           "toc" = rep(c("1sym", "2asym_l", "2asym_r"), each = length(x))) %>%
  ggplot(aes(x = x, y = y, col = toc)) +
  geom_line(aes(linetype = toc),size = 1.5, alpha = 0.6) +
  geom_segment(aes(x = 2000, xend = 2000, y = 0, yend = 0.9),
               col = "grey20",
               lty = 2) +
  ylim(c(0, 1)) +
  scale_color_manual(values = c("grey40", "red", "darkgreen"),
                     labels = c("Symmetric", "Left-skewed", "Right-skewed")) +
  scale_linetype_manual(values = c(1, 6, 6),
                        labels = c("Symmetric", "Left-skewed", "Right-skewed")) +
  labs(color = "Type of SRC",
       linetype = "Type of SRC") +
  theme_bw()
```

Fig S4.1: Three types of species response curves sharing a same optimum (dashed vertical segment).

### Simulated datasets

We defined three potential shapes of species response curves: symmetric, asymmetric with a longer left tail and asymmetric with a longer right tail (Fig. S4.1). We simulated 9 species to describe all possible combinations of SRC that could occur when two sampling occasions are considered (Fig. S4.2). To keep it simple we chose to only simulate specialist species, shifting upward and having their optima of both sampling occasions in the middle of the sampling range. We simulated curve widths ranged from 400 m to 600 m. Maximum occupancy probabilities were obtained by random draws from a uniform distribution bounded between 0.85 and 0.99. Species optima were randomly positioned between 1490 m and 2460 m, in the middle range of the unbalanced sampling design (i.e. sampling scenario A2 of the manuscript). We assigned only upward shifts following random draws from a uniform distribution bounded between 80 and 120 m (Fig. S4.2).

```{r simulated curves, fig.height=10, fig.width=8}
curve.type1 <- rep(c("Sym", #symmetric
                        "R-skd", #asymmetric to the right
                        "L-skd"), #asymmetric to the left
                      3)

curve.type2 <- rep(c("Sym", #symmetric
                        "R-skd", #asymmetric to the right
                        "L-skd"), #asymmetric to the left
                      each = 3)

par(mfrow = c(3,3))
for (i in 1:N){
  plot(X_A1[order(X_A1)], psi[,i,1,1], 
       type = "l", 
       xlab = "", ylab = "", 
       ylim = c(0,1.05),
       xlim = range(X_A1),
       col = "blue",
       main = paste(curve.type1[i],
                    curve.type2[i]))
  segments(x0 = data.sp[i, ]$opt,
           x1 = data.sp[i, ]$opt,
           y0 = - 1, 
           y1 = data.sp[i, ]$max.occ1,
           lty = 2,
           col = "blue")
  lines(X_A1[order(X_A1)], psi[,i,2,1],
        col = "red")
  segments(x0 = (data.sp[i, ]$opt + data.sp[i, ]$shift),
           x1 = (data.sp[i, ]$opt + data.sp[i, ]$shift),
           y0 = - 1, 
           y1 = data.sp[i, ]$max.occ2,
           lty = 2,
           col = "red")
  text(x = (data.sp[i, ]$opt + data.sp[i, ]$shift / 2),
       y = 1,
       labels = paste("+", round(data.sp[i, ]$shift, 0), "m"))
  arrows(x0 = (data.sp[i, ]$opt),
         x1 = (data.sp[i, ]$opt + data.sp[i, ]$shift),
         y0 = 0.95, 
         y1 = 0.95,
         length = 0.05)
}
```

Fig S4.2: Species response curves of the nine simulated species. Blue curves represent the SRCs of the first sampling occasion. Panel titles indicates the shapes of the two SRCs, with sym, r_skd and l_skd standing for symmetric, right-skewed and left-skewed.

Species response curves were simulated using a mixture of two normal distributions with a common mean and two variance parameters (see Eq. (1) bellow). The common mean represents the species optimum. The two variances are related to the width of the SRC at each side of the species optimum. The first distribution (resp. second) is used to simulate the left (resp. right) side of the optimum of the SRC. Hence, if the first normal distribution has a greater standard deviation than the second normal distribution, the curve will be left-skewed. If both distribution have the same standard deviation, the SRC is symmetric. Note that we used the density function of the normal distribution multiplied by a scaling parameter to reach a specified probability. The scaling parameter is defined as the ratio of the maximum occupancy probability intended and the maximum density of the normal distribution. We simulated the site-specific probability of occupancy for each species as follows:

```{=tex}
\begin{equation}
    \psi^s_{i,j,k} =
    \begin{cases}
      a_i \frac{exp( - \frac{(x^s_j - (\theta_i + (S_k \times \delta_i)))^2}{2 \sigma_1i ^ 2} )}{\sigma_1i \sqrt{2 \pi}} & \text{for $x^s_j \leq \theta_i$}\\
      b_i \frac{exp( - \frac{(x^s_j - \theta_i + (S_k \times \delta_i))^2}{2 \sigma_2i ^ 2} )}{\sigma_2i \sqrt{2 \pi}} & \text{for $x^s_j > \theta_i$}
    \end{cases}
\end{equation}
```
with $a_i$ and $b_i$ the scaling parameters related to the maximum probability, $\theta$ the species optimum, $\delta$ the species shift and $\sigma$ the standard deviations related to the width.

Once SRCs were simulated for each species and sampling occasion, we generated presence/absence data, $Y^s_{i,j,k}$, from random draws of a Bernoulli distribution: $Y^s_{i,j,k} \sim \text{Bernoulli}(\psi^s_{i,j,k})$. We replicated the process 30 times for the two sampling scenarios. Contrary to the manuscript simulation design, here the simulated parameters of the SRCs ($a_i$, $\theta_i$, $\delta_i$, $\sigma_i$) were identical for the two sampling scenarios.

### Accuracy of optimum shift estimates under violiation of the symmetry assumption

In our simulations, all three methods respond in the same way to departure from assumption of symmetric unimodal species-gradient relationships (Fig. S4.3, Table S4.1). More than the departure of symmetry assumption, the models are sensitive to changes in SRC shapes between the two sampling occasions. The more difference between SRC shapes, the larger the bias for all methods. For instance, the highest bias were obtained for species having a SRC skewed toward one side of the gradient at the first sampling occasion and a SRC skewed toward the other side in the next sample (e.g. left-skewed to right-skewed). In contrast, methods produced accurate optimum shift estimates for species having same skewness during the two sampling occasions (e.g. species 5).

```{r include=FALSE}
full.res.shift %>%
  dplyr::mutate("abs.bias" = abs(estim.shift - true.shift)) %>%
  dplyr::group_by(model, scenario) %>%
  dplyr::summarise_at("abs.bias", c("mean" = mean, "sd" = sd)) -> summary_abs.bias

full.res.shift %>%
  dplyr::mutate("bias" = estim.shift - true.shift) %>%
  dplyr::filter(sp %in% paste0("sp", c(1, 5, 9))) %>%
  dplyr::group_by(model, scenario) %>%
  dplyr::summarise_at("bias", c("mean" = mean, "sd" = sd))

summary_bias %>%
  dplyr::mutate("mean" = abs(mean)) %>%
  dplyr::group_by(model, scenario) %>%
  dplyr::summarise_at("mean", mean)

summary_bias %>%
  dplyr::filter(sp %in% paste0("sp", c(2, 3, 4, 6, 7, 8))) %>%
  dplyr::mutate("mean" = abs(mean)) %>%
  dplyr::group_by(model, scenario) %>%
  dplyr::summarise_at("mean", mean)
```

```{r bias table}
summary_bias %>%
  dplyr::ungroup() %>%
  dplyr::inner_join(full.res.shift[, c("sp", "true.shift")], by = "sp") %>%
  dplyr::distinct() %>%
  dplyr::mutate("true.shift" = round(true.shift),
                "bias" = paste0(round(mean), " (", round(sd), ")")) %>%
  dplyr::select(-c('mean', "sd")) %>%
  tidyr::pivot_wider(names_from = model, values_from = bias) %>%
  dplyr::mutate("SRC 1" = rep(rep(c("Symmetric", 
                                 "Right-skewed", 
                                 "Left-skewed"), 
                               3), 2),
                "SRC 2" = rep(rep(c("Symmetric", 
                                 "Right-skewed", 
                                 "Left-skewed"), 
                               each = 3), 2)) %>%
  dplyr::rename("True shift" = "true.shift", 
                "EHMOS" = "ehmos", 
                "cGLMM" = "glmm", 
                "t-test" = "t.test")  %>%
  dplyr::select(c("SRC 1", "SRC 2", "True shift", "EHMOS", "cGLMM", "t-test")) %>%
  kableExtra::kable(format = "latex",
                    caption = "Average of bias in optimum shift estimates for the two simulated scenarios after 30 replications. Numbers in brackets represent standard deviations. ",
                    booktabs = TRUE,
                    longtable = TRUE) %>%
  kableExtra::kable_styling(latex_options = "HOLD_position") %>%
  kableExtra::pack_rows("Sampling scenario A2", 1, 9) %>% # inversion of scenario names during data simulation
  kableExtra::pack_rows("Sampling scenario A1", 10, 18)
```

```{r bias figure}
full.res.shift %>%
  dplyr::mutate("bias" = (estim.shift - true.shift)) %>%
  ggplot(aes(x = sp, y = bias)) +
  geom_point(color = alpha("grey20", 0.4)) +
  facet_grid(scenario ~ model,
             labeller = labeller(.rows = c("A1" = "A2", "A2" = "A1"))) + # correct past error of confounding names when creating data
  geom_point(data = summary_bias,
             aes(x = sp, y = mean), 
             color = "red",
             size = 2) +
  geom_segment(data = summary_bias,
               aes(x = sp, 
                   xend = sp, 
                   y = mean - sd,
                   yend = mean + sd), 
               color = "red") +
  ylab("Bias (m)") +
  scale_x_discrete(labels = paste(curve.type1,
                    curve.type2)) +
  theme_bw() +
  theme(axis.text.x = element_text(vjust = 1,
                                   hjust = 1,
                                   angle = 45),
        axis.title.x = element_blank())
```

Figure S4.3: Bias in optimum shift estimates with simulated species having different kinds of unimodal species response curves (symmetric, right-skewed or left-skewed). Grey points correspond to results from one replication. Mean bias are represented with their standard deviations by red points and red segments.

### Detection of departure from model assumption

We tried two approaches to test the departure from the symmetry assumption made by the three presented models: 1) a Bayesian P-value approach that is often used by ecologists to detect a lack-of-fit in Bayesian models [@conn2018], and 2) a HOF approach previously used to test for the symmetry assumption in range shift studies [@wilson2005].

#### Goodness-of-fit

We assess quality of model fit using a Bayesian P-value approach [@gelman2013] for each species. We used the discrepancy measure defined by @zipkin2009: $D(y_i) = \sum_{k}(y_{ik} - \hat\psi_{ik})^2$ for the presence-absence observations $y_{ik}$ of species $i$ and their expected values $\hat\psi_{ik}$ under the model at MCMC replication $k$. This discrepancy statistic is computed independently for each species $i$. A reference distribution is computed by simulating data sets from the posterior distribution, $y_{ik}^{sim} \sim p(\theta_{ik} | y)$, and computing the discrepancy measure, $D(y_{ik}^{sim})$, for the simulated data sets. The Bayesian p-value for species $i$ is defined as the probability: $p_{Bi} = \text{Pr}(D(y_i)>D(y_i^{sim}))$. Extreme values (e.g. less than 0.05 or greater than 0.95) are interpreted as a lack of fit.

```{r table GoF}
data.frame("SRC 1" = rep(c("Symmetric", 
                                 "Right-skewed", 
                                 "Left-skewed"), 
                               3),
                "SRC 2" = rep(c("Symmetric", 
                                 "Right-skewed", 
                                 "Left-skewed"), 
                               each = 3),
           "cGLMM" = apply(bp.sp_glmm, 1, function(x)(sum(x < 0.05 | x > 0.95))),
           "EHMOS" = apply(bp.sp_ehmos, 1, function(x)(sum(x < 0.05 | x > 0.95)))) %>%
  kableExtra::kable(format = "latex",
                    caption = "Number of times where lack-of-fit were detected after 30 replications for the 9 species simulated with different shapes of response.",
                    booktabs = TRUE,
                    longtable = TRUE) %>%
  kableExtra::kable_styling(latex_options = "HOLD_position")
```

We detected no lack-of-fit according to the Bayesian p-value approach, except for one species in one replication of the simulation (Table S4.2).

Bayesian p-value approach seems conservative in regards of skewed species response. Thus, it should not be used to investigate departure from model assumption on shape of species response.

#### HOF approach

HOF approach consists in applying seven models and select the model that best fit the data [@huisman1993; @jansen2013]. The five models describe five shapes of response curves assumed to be ecologically meaningful [@oksanen2002; @jansen2013]: I) flat, II) monotone, III) plateau, IV) unimodal symmetric, V) unimodal skewed, VI) bimodal symmetric and VII) bimodal skewed. This approach has been used to determine the shape of species response along environmental gradients in many studies [e.g. @oksanen2002; @jansen2013; @michaelis2017], and especially in range shift studies [@urli2014, @wilson2005]. For instance, @wilson2005 used the HOF approach to verify the assumption of symmetrical species response curves made by their two-GLMs approach. We tested this method on our simulated data sets. We applied the HOF approach on each replicated data set and each species for sampling occasion one using the `HOF` function of the R-package `eHOF` [@jansen2013].

```{r}
shape.hat %>% 
  as.data.frame() %>% 
  dplyr::mutate_all(as.factor) %>% 
  tidyr::pivot_longer(cols = 1:9,
                      names_to = "species",
                      values_to = "hof_shape") %>%
  dplyr::group_by(species) %>%
  dplyr::count(hof_shape) %>%
  tidyr::pivot_wider(names_from = "hof_shape",
                     values_from = "n") %>%
  dplyr::ungroup() -> hof_res

hof_res[is.na(hof_res)] <- 0

hof_res %>%
  dplyr::mutate("species" = 1:9,
                "true shape" = rep(c("Symmetric (IV)", 
                                 "Right-skewed (V)", 
                                 "Left-skewed (V)"), 
                               3)) %>%
  dplyr::select(1, 5, 2:4) %>%
  kableExtra::kable(format = "latex",
                    caption = "Model choices of the HOF approach for the 9 simulated species after 30 replications.",
                    booktabs = TRUE,
                    longtable = TRUE) %>%
  kableExtra::add_header_above(c(" " = 2, "HOF models" = 3))
```

The HOF approach failed to retrieve the correct shape for numerous species and iterations (Table S4. 3). It estimated symmetrical response in majority of replications even for species having skewed response. Species with symmetrical responses had been misclassified in at least 15\% of the replications.

### Discussion

We found that departure from assumption of symmetry in species response could induce bias in optimum shift estimates of the three methods. The estimates were biased when degree of skewness in species response changed between the two sampling occasions. Thus, readers should be cautious regarding the shape of species response curves, and more importantly in the potential changes in shapes of SRC between the samples studied, before applying any of the three methods presented here.

We tried two approaches in order to detect departure from model assumption of symmetry in SRC: 1) Bayesian p-value approach as a measure of goodness-of-fit [@gelman2013] and 2) HOF approach that were used by @wilson2005 in this context. None was found to be a reliable evaluation of the symmetry assumption based on our simulations. This could be due to the relatively low degree of skewness of the simulated SRCs, but previous works have also shown that HOF approach could misclassify SRC shapes even with high degree of skewness [@jansen2013; @michaelis2017]. @jansen2013 found for instance that HOF approach could be sensitive to optimum position along the environmental gradient. Their results indicate numerous misclassifications of HOF approach (see Table 4 and Fig. A3 in @jansen2013), especially (but not only) for species having their optimum close to sampling limits (i.e. edge species). These results also stand for GAMs [@jansen2013; @citores2020]. Besides, @oksanen2002 pointed out that defining symmetry or asymmetry with GAMs could be subjective and thus advice to use of HOF approach. For these reasons we do not used GAMs for testing the symmetry assumption.

### Conlusion

It appears important to thoroughly inspect SRC shape before analysis due to potential bias when SRC shapes change over sampling occasions. More researches are needed to be able to test for asymmetrical SRC, especially simulation studies. Ecologists could also rely on more precise data, e.g. abundance data, when they are available to test the assumption of symmetrical speceis response curves.

### References

Citores L., Ibaibarriaga L., Lee D.-J., Brewer M.J., Santos M., & Chust G. (2020) Modelling species presence–absence in the ecological niche theory framework using shape-constrained generalized additive models. _Ecological Modelling_, **418**, 108926.

Conn P.B., Johnson D.S., Williams P.J., Melin S.R., & Hooten M.B. (2018) A guide to bayesian model checking for ecologists. _Ecological Monographs_, **88**, 526–542.

Gelman A., Carlin J.B., Stern H.S., Dunson D.B., Vehtari A., & Rubin D.B. (2013) _Bayesian Data Analysis_. Chapman; Hall, New York.

Huisman J., Olff H., & Fresco L. (1993) A hierarchical set of models for species response analysis. _Journal of Vegetation Science_, **4**, 37–46.

Jansen F. & Oksanen J. (2013) How to model species responses along ecological gradients–huisman–olff–f resco models revisited. _Journal of Vegetation Science_, **24**, 1108–1117.

Michaelis J. & Diekmann M.R. (2017) Biased niches – Species response curves and niche attributes from Huisman-Olff-Fresco models change with differing species prevalence and frequency. _PLOS ONE_, **12**, e0183152.

Oksanen J. & Minchin P.R. (2002) Continuum theory revisited: What shape are species responses along ecological gradients? _Ecological Modelling_, **157**, 119–129.

Urli M., Delzon S., Eyermann A., Couallier V., García-Valdés R., Zavala M.A., & Porté A.J. (2014) Inferring shifts in tree species distribution using asymmetric distribution curves: A case study in the Iberian mountains. _Journal of Vegetation Science_, **25**,
147–159.

Wilson R.J., Gutiérrez D., Gutiérrez J., Martínez D., Agudo R., & Monserrat V.J. (2005) Changes to the elevational limits and extent of species ranges associated with climate change. _Ecology letters_, **8**, 1138–1146.

Zipkin E.F., DeWan A., & Royle J.A. (2009) Impacts of forest fragmentation on species richness: A hierarchical approach to community modelling. _Journal of Applied Ecology_, **46**, 815–822.
